---
title: "Logarithmic axes with linear tickmarks in a single R plot"
author: "Petr Keil"
date: "July 6, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

I have always loved [Mathematica's](http://media.virtual-maxim.de/uploads/2010/12/ErrorBarLogPlots_Mathmatica.png) and [Matlab's log-log plots](http://www.mathworks.com/matlabcentral/mlc-downloads/downloads/submissions/35265/versions/2/previews/html/Loglog_Plot_01.png) with logarithmic axes and linear
tickmarks. In a way, they enable to visualize both multiplication and addition in a single image. They also enable to more easily assign values to data points.

I haven't found a simple 'one-liner' way to do such plots in R. Actually, I have always found R's treatment of logarithmic axes a bit dull and unsatisfactory.

# Function `loglog.plot`

To provide the log-linear plotting functionality, I have made my own function `loglog.plot`.

**Arguments:**

`xlim, ylim` - Numeric vectors of length 2, giving the x and y coordinates ranges.

`xlog, ylog` - Logical value indicating if x and y axes should be logarithmic or linear. In case they are linear, no grid is drawn.  

`xbase, ybase` - Base of the logarithm of the respective axes.

`...` - Further arguments to the generic R function `plot`.

**The function:**

```{r}
source("loglogplot.r")
```

**Example 1 - both axes logarithmic**

```{r, fig.width=8, fig.height=8}
  loglog.plot(xlab="log with base of 10", ylab="log with base of 2")
```

**Example 2 - x axis linear, y axis logarithmic**

```{r, fig.width=8, fig.height=8}
  loglog.plot(xlab="log with base of 10", ylab="log with base of 2")
```