---
title: "Logarithmic axes with linear tickmarks in a single R plot"
author: "Petr Keil"
date: "July 6, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

I have always loved [Mathematica's](http://media.virtual-maxim.de/uploads/2010/12/ErrorBarLogPlots_Mathmatica.png) and [Matlab's log-log plots](http://www.mathworks.com/matlabcentral/mlc-downloads/downloads/submissions/35265/versions/2/previews/html/Loglog_Plot_01.png) with logarithmic axes and linear
tickmarks. In a way, they enable to visualize both multiplication and addition in a single image. They also enable to more easily assign values to data points.

I haven't found a simple 'one-liner' way to do such plots in R. Actually, I have always found R's treatment of logarithmic axes a bit dull and unsatisfactory.

# Function `loglog.plot`

To provide the log-linear plotting functionality, I have made my own function `loglog.plot`.

**Arguments:**

`xlim, ylim` - Numeric vectors of length 2, giving the x and y coordinates ranges.

`xlog, ylog` - Logical value indicating if x and y axes should be logarithmic or linear. In case they are linear, no grid is drawn.  

`xbase, ybase` - Base of the logarithm of the respective axes.

`...` - Further arguments to the generic R function `plot`.

**The function:**

```{r}
loglog.plot <- function(xlim=c(0.01,1000), xlog=TRUE, xbase=10, 
                        ylim=c(0.3,1000), ylog=TRUE, ybase=2, ...)
{
  if(xlog) # rounding the X-axis limits on the log scale
  {
    xlim[1] <- floor(log(xlim[1], base=xbase))
    xlim[2] <- round(log(xlim[2], base=xbase))
    xbreaks <- xlim[1]:xlim[2]
  }
  if(ylog) # rounding the Y-axis limits on the log scale
  { 
    ylim[1] <- floor(log(ylim[1], base=ybase))
    ylim[2] <- round(log(ylim[2], base=ybase))
    ybreaks <- ylim[1]:ylim[2]
  }

  ### the empty plot into which the axes will be drawn ###
  plot(xlim, ylim, type="n", axes=FALSE, frame=TRUE, ...)

  if(xlog) # plotting the X-axis tickmarks and grids
  {
    for(x in xbase^xbreaks)
    {
      subx <- log(seq(from=x, to=x*xbase, length=xbase) , base=xbase )
      abline(v=subx, col="grey")
    }
    axis(side=1, at=xbreaks, labels=xbase^xbreaks, tck=0.02)
  }
  else  axis(side=1, tck=0.02)

  if(ylog) # plotting the Y-axis tickmarks and grids
  {
    for(y in ybase^ybreaks)
    {
      suby <- log(seq(from=y, to=y*ybase, length=ybase), base=ybase)  
      abline(h=suby, col="grey")
    }
    axis(side=2, at=ybreaks, labels=ybase^ybreaks, las=2, tck=0.02)
  }
  else axis(side=1, tck=0.02)
}
```

**Example 1 - both axes logarithmic**

```{r, fig.width=8, fig.height=8}
  loglog.plot(xlab="log with base of 10", ylab="log with base of 2")
```

**Example 2 - x axis linear, y axis logarithmic**

```{r, fig.width=8, fig.height=8}
  loglog.plot(xlab="log with base of 10", ylab="log with base of 2")
```